---
title: "Vermont"
---
<style type="text/css">
h1.title {
  text-align: center;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
# library(flexdashboard)
library(tidyverse)
library(sf)
library(tmap)
library(maptools)
# library(janitor)
# library(kableExtra)
library(tigris)
options(tigris_use_cache = TRUE, tigris_class = "sf")
library(leaflet)
library(leaflet.extras)
library(DT)
# set common data table options
# options(DT.options = list(scrollY="100vh", lengthMenu = c(5, 10, 15, 20)))
options(DT.options = list(lengthMenu = c(10, 20, 50, 100)))
library(highcharter)
library(rmapshaper)
library(tidycensus)
```

```{r data, include=FALSE, cache=TRUE}
load("../DATA/ne_layers.rds")

# isolate MA census geography and relevant variables
vt_blkgrps_sf <- ne_blkgrp_sf %>% 
  filter(STATE == "Vermont") %>% 
  select(GEOID, NAME, STATE, bg_area_m2, totalpopE, minorityE, minority_pctE, under5E, pct_under5E, under18E, pct_under18E, over64E, pct_over64E, householdsE, eng_hhE, eng_limitE, eng_limit_pctE, age25upE, lthsE, pct_lthsE, povknownE, num2povE, pct2povE) %>%
  st_transform(., crs = 4326) %>% 
  filter(!st_is_empty(.)) 

vt_tracts_sf <- ne_tracts_sf %>% 
  filter(STATE == "Vermont") %>% 
  select(GEOID, NAME, totalpopE, STATE, Over18E, disabledOver18E, 
         pct_disabilityOver18E, totalHHE, HHnoCarE, pct_HHnoCarE) %>% 
  st_transform(., crs = 4326) %>% 
  filter(!st_is_empty(.))

# # clean up
# rm(list = ls(pattern = "ne_"))

# Read in evacuation risk data. Developed parts of census units; need to filter and join.
load("../DATA/FEMA/VT/nfhza_census.Rds")

# join total flood records where percentage of pop of concern in flood or river corridor prone areas is 80th percentile
vt_blkgrps_sf <- vt_blkgrps_flood %>% 
  as.data.frame() %>% 
  mutate(NewEngHH = Proportion*eng_hhE,
         NewAge25Up = Proportion*age25upE) %>%
  group_by(GEOID) %>% 
  summarize(across(NewPop:NewAge25Up, ~ sum(.x, na.rm = T), 
                     .names = "sum_{.col}")) %>% 
  filter(percent_rank(sum_NewMinority/sum_NewPop) >= 0.8 | 
           percent_rank(sum_NewUnder5/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewOver64/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewUnder18/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewEng_limit/sum_NewEngHH) >= 0.8 |
           percent_rank(sum_NewPov/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewLths/sum_NewAge25Up) >= 0.8) %>% 
  transmute(GEOID = GEOID, floodRisk = "F") %>% 
  left_join(vt_blkgrps_sf,., by="GEOID")


# Re-process FEMA and RC only affected block groups
library(parallel)
##### calculate areas at risk from FEMA flood zones and River Corridors separately for map labeling #####
# load cropped flood zones
vt_nfhza_2852_land <- readRDS("../DATA/FEMA/VT/vt_nfhza_2852_land.Rds") %>% 
  mutate(Area = as.numeric(Area))

# read in river corridors layer. downloaded from VT Open Geodata Portal (https://geodata.vermont.gov/datasets/VTANR::river-corridors-august-2019) in December 2020. 
rc <- st_read(dsn = "../DATA/FEMA/VT/WaterHydro_RiverCorridors",
              layer = "WaterHydro_RiverCorridors_poly") %>% 
  st_transform(., crs = 2852) %>% 
  st_make_valid()

# read in developed block group polygons
vt_blkgrps_developed <- st_read(dsn = "../DATA/FEMA/VT",
                                layer = "vt_blkgrps_developed") %>% 
  st_transform(., crs = 2852) %>%
  st_make_valid()

# calculate populations within flood risk areas
# USE PARALLEL PROCESSING
# determine number of cores to use for parallel processing
n.cores <- detectCores()-1

# create parallel cluster nodes
clust <- makeCluster(n.cores)

# use ClusterEvalQ to load needed packages in each cluster
clusterEvalQ(clust, {
  library(tidyverse)
  library(sf)
  library(tmaptools)
})

# export variables to each node for use in processing
clusterExport(clust, varlist = c("vt_nfhza_2852_land", "vt_blkgrps_developed"))

# break up data by county for faster processing
# download counties layer
vt_counties <- counties(state = "VT", cb = TRUE) %>% 
  st_transform(.,crs=2852)

# break up counties into a list to run with parLapply
county_list <- list()
for (i in 1:(nrow(vt_counties))){
  county_list[[i]] <- vt_counties[i,]
}

# run in parallel
vt_blkgrps_nfhza_list <- parLapply(clust, county_list, function(x){
  crop_shape(vt_nfhza_2852_land, x, polygon = T) %>% 
    st_intersection(vt_blkgrps_developed, .) %>% 
    mutate(NewArea = as.numeric(st_area(.)))
} )

stopCluster(clust)

# bring it all together
vt_blkgrps_nfhza <- do.call(rbind, vt_blkgrps_nfhza_list) %>% 
  st_make_valid()

# allocate populations to intersected areas
vt_blkgrps_nfhza <- vt_blkgrps_nfhza %>% 
  left_join(., st_drop_geometry(vt_blkgrps_sf), by = "GEOID") %>% 
  filter(!st_is_empty(.)) %>% 
  mutate(Proportion = as.numeric(NewArea/OldArea),
         NewPop = totalpopE*Proportion,
         NewMinority = minorityE*Proportion,
         NewUnder5 = under5E*Proportion,
         NewOver64 = over64E*Proportion,
         NewUnder18 = under18E*Proportion,
         NewEng_limit = eng_limitE*Proportion,
         NewPov = num2povE*Proportion,
         NewLths = lthsE*Proportion)


# repeat for river corridors
# create parallel cluster nodes
clust <- makeCluster(n.cores)

# use ClusterEvalQ to load needed packages in each cluster
clusterEvalQ(clust, {
  library(tidyverse)
  library(sf)
  library(tmaptools)
})

# export variables to each node for use in processing
clusterExport(clust, varlist = c("rc", "vt_blkgrps_developed"))

# run in parallel
vt_blkgrps_rc_list <- parLapply(clust, county_list, function(x){
  crop_shape(rc, x, polygon = T) %>% 
    st_intersection(vt_blkgrps_developed, .) %>% 
    mutate(NewArea = as.numeric(st_area(.)))
} )

stopCluster(clust)

# bring it all together
vt_blkgrps_rc <- do.call(rbind, vt_blkgrps_rc_list) %>% 
  st_make_valid()

# allocate populations to intersected areas
vt_blkgrps_rc <- vt_blkgrps_rc %>% 
  left_join(., st_drop_geometry(vt_blkgrps_sf), by = "GEOID") %>% 
  filter(!st_is_empty(.)) %>% 
  mutate(Proportion = as.numeric(NewArea/OldArea),
         NewPop = totalpopE*Proportion,
         NewMinority = minorityE*Proportion,
         NewUnder5 = under5E*Proportion,
         NewOver64 = over64E*Proportion,
         NewUnder18 = under18E*Proportion,
         NewEng_limit = eng_limitE*Proportion,
         NewPov = num2povE*Proportion,
         NewLths = lthsE*Proportion)
#########

# Add labels to identify NFHZA prone areas separately from river corridors
vt_blkgrps_sf <- vt_blkgrps_nfhza %>% 
  as.data.frame() %>% 
  mutate(Proportion = NewArea/OldArea,
         NewEngHH = Proportion*eng_hhE,
         NewAge25Up = Proportion*age25upE,
         floodtype = if_else(FLD_ZONE == "X", "500-year", "100-year")) %>%
  group_by(GEOID) %>% 
  summarize(across(NewPop:NewAge25Up, ~ sum(.x, na.rm = T), 
                   .names = "sum_{.col}"),
            floodtypes = paste(unique(floodtype), collapse = ", ")) %>% 
  filter(percent_rank(sum_NewMinority/sum_NewPop) >= 0.8 | 
           percent_rank(sum_NewUnder5/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewOver64/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewUnder18/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewEng_limit/sum_NewEngHH) >= 0.8 |
           percent_rank(sum_NewPov/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewLths/sum_NewAge25Up) >= 0.8) %>% 
  transmute(GEOID = GEOID, nfhzaRisk = "F", floodtypes = floodtypes) %>% 
  left_join(vt_blkgrps_sf,., by="GEOID")

# Add labels to identify river corridor risk areas separately from FEMA
vt_blkgrps_sf <- vt_blkgrps_rc %>% 
  as.data.frame() %>% 
  mutate(Proportion = NewArea/OldArea,
         NewEngHH = Proportion*eng_hhE,
         NewAge25Up = Proportion*age25upE,
         floodtype = " River Corridor") %>%
  group_by(GEOID) %>% 
  summarize(across(NewPop:NewAge25Up, ~ sum(.x, na.rm = T), 
                   .names = "sum_{.col}"),
            floodtype = unique(floodtype)) %>% 
  filter(percent_rank(sum_NewMinority/sum_NewPop) >= 0.8 | 
           percent_rank(sum_NewUnder5/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewOver64/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewUnder18/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewEng_limit/sum_NewEngHH) >= 0.8 |
           percent_rank(sum_NewPov/sum_NewPop) >= 0.8 |
           percent_rank(sum_NewLths/sum_NewAge25Up) >= 0.8) %>% 
  transmute(GEOID = GEOID, rcRisk = "RC", floodtype = floodtype) %>% 
  left_join(vt_blkgrps_sf,., by="GEOID")


#create layer of municipalities
vt_towns <- county_subdivisions("VT") %>% 
  filter(NAME != "County subdivisions not defined") %>%  
  st_transform(., crs = 4326) %>% 
  select(GEOID, NAME, NAMELSAD)

# Assign municipality names to block groups
vt_blkgrps_sf <- county_subdivisions("VT") %>% 
  st_transform(., crs = st_crs(vt_blkgrps_sf)) %>% 
  transmute(TOWN = NAMELSAD) %>% 
  st_join(vt_blkgrps_sf, ., largest = TRUE) %>% 
  mutate(NAME = str_remove_all(NAME, ", Vermont")) %>% 
  st_transform(., crs = 4326)

# create layer of state house districts
house_districts <- st_read(
  "../DATA/shapefiles/VT_Data_-_Vermont_House_Districts_2012-shp",
  "VT_Data_-_Vermont_House_Districts_2012") %>% 
  transmute(REP_DIST = NAME) %>% 
  st_transform(., crs = 4326) %>% 
  st_make_valid()

# create layer of state senate districts
senate_districts <- st_read("../DATA/shapefiles/VT_Senate_Districts_2012-shp",
                            "VT_Senate_Districts_2012") %>% 
  transmute(SEN_DIST = NAME) %>% 
  st_transform(., crs = 4326) %>% 
  st_make_valid()

# Assign state house district names to block groups
vt_blkgrps_sf <- house_districts %>% 
  select(REP_DIST) %>% 
  st_join(vt_blkgrps_sf, ., largest = TRUE)

# Assign state senate district names to block groups
vt_blkgrps_sf <- senate_districts %>% 
  select(SEN_DIST) %>% 
  st_join(vt_blkgrps_sf, ., largest = TRUE)


# Create labeling variables to identify pops of a concern in a given block group that contributed to cumulative burden score
vt_blkgrps_sf <- vt_blkgrps_sf %>% 
  mutate(Minority80th = if_else(percent_rank(minority_pctE) >= 0.8, "People of Color","NA"),
    Under5_80th = if_else(percent_rank(pct_under5E) >= 0.8, "Under 5", "NA"),
    Under18_80th = if_else(percent_rank(pct_under18E) >= 0.8, "Under 18", "NA"),
    Over64_80th = if_else(percent_rank(pct_over64E) >= 0.8, "Over 64", "NA"),
    lths80th = if_else(percent_rank(pct_lthsE) >= 0.8, "No HS Diploma", "NA"),
    pct2pov80th = if_else(percent_rank(pct2povE) >= 0.8, "Low Income", "NA"),
    eng_limit_pct80th = if_else(percent_rank(eng_limit_pctE) >= 0.8, "Limited English HH", "NA"),
    POPSlabel = gsub("^,*|(?<=,),|,*$", "", # get rid of extra commas
    str_remove_all( # get rid of NAs
      paste(Minority80th,
            Under5_80th,
            Under18_80th,
            Over64_80th,
            lths80th,
            pct2pov80th,
            eng_limit_pct80th, sep = ","),
      pattern = "NA"), 
    perl=T
    ),
    POPSlabel = if_else(POPSlabel == "", "No Pops of Concern", POPSlabel)
    )
  # filter(POPSlabel != "No Pops of Concern")


# Calculate Pop and Pct pop at risk by jurisdiction
NFHZAbyTown <- vt_blkgrps_flood %>% 
  transmute(NewPop = NewPop, OldArea = as.numeric(st_area(.))) %>% 
  st_intersection(st_transform(vt_towns, st_crs(vt_blkgrps_flood))) %>% 
  mutate(NewArea = as.numeric(st_area(.)),
         Proportion = NewArea/OldArea,
         nfhzaPop = as.integer(NewPop*Proportion)) %>% 
  as.data.frame() %>% 
  group_by(NAMELSAD) %>% 
  summarize(nfhzaPop = sum(nfhzaPop))

# download town pops from tidycensus
town_pops <- get_acs(geography = "county subdivision", survey = "acs5",
                     variables = c(totalpop = "B03002_001"),
                     state = "VT", output = "wide", year = 2018) %>% 
  select(GEOID, totalpopE)

# # grab municipal boundaries from TIGRIS
# vt_towns_sf <- county_subdivisions(state = "VT", cb = TRUE) %>% 
#   st_transform(., crs = 26986)

# create df with town names from tigris and pops from tidycensus
town_nfhzaPop_pct <- vt_towns %>% 
  as.data.frame() %>% 
  left_join(., town_pops, by = "GEOID") %>% 
  select(NAMELSAD, totalpopE) %>% 
  left_join(., NFHZAbyTown,by = "NAMELSAD") %>% 
  replace_na(list(nfhzaPop = 0)) %>% 
  mutate(PctPopNFHZA = if_else(totalpopE == 0, 0, nfhzaPop/totalpopE*100))


# calculate flood pops by house district
# create pop totals for each district
house_names_pops <- get_acs(geography = "block group", survey = "acs5",
                     variables = c(totalpop = "B03002_001"), state = "VT", 
                     output = "wide", year = 2018, geometry = TRUE) %>% 
  select(GEOID, totalpopE) %>% 
  st_transform(.,st_crs(vt_blkgrps_flood)) %>% 
  select(totalpopE) %>% 
  mutate(OldArea = as.numeric(st_area(.))) %>% 
    st_intersection(
      st_make_valid(
        st_transform(house_districts,st_crs(vt_blkgrps_flood))
        )
      ) %>% 
  mutate(NewArea = as.numeric(st_area(.)),
         Proportion = NewArea/OldArea,
         NewPop = totalpopE*Proportion) %>% 
  as.data.frame() %>% 
  group_by(REP_DIST) %>% 
  summarize(totalpopE = sum(NewPop))

# identify FEMA risk pops by district and calculate pop pct
NFHZAbyHouse <- vt_blkgrps_flood %>% 
  transmute(NewPop = NewPop, OldArea = as.numeric(st_area(.))) %>% 
  st_intersection(
    st_make_valid(
      st_transform(house_districts, st_crs(vt_blkgrps_flood))
      )
    ) %>% 
  mutate(NewArea = as.numeric(st_area(.)),
         Proportion = NewArea/OldArea,
         nfhzaPop = as.integer(NewPop*Proportion)) %>% 
  as.data.frame() %>% 
  group_by(REP_DIST) %>% 
  summarize(nfhzaPop = sum(nfhzaPop)) %>% 
  left_join(house_names_pops, ., by = "REP_DIST") %>% 
  replace_na(list(nfhzaPop = 0)) %>% 
  mutate(PctPopNFHZA = nfhzaPop/totalpopE*100)


# calculate flood pops by senate district
# create pop totals for each district
senate_names_pops <- get_acs(geography = "block group", survey = "acs5",
                     variables = c(totalpop = "B03002_001"), state = "VT", 
                     output = "wide", year = 2018, geometry = TRUE) %>% 
  select(GEOID, totalpopE) %>% 
  st_transform(.,st_crs(vt_blkgrps_flood)) %>% 
  select(totalpopE) %>% 
  mutate(OldArea = as.numeric(st_area(.))) %>% 
    st_intersection(
      st_make_valid(
        st_transform(senate_districts,st_crs(vt_blkgrps_flood))
        )
      ) %>% 
  mutate(NewArea = as.numeric(st_area(.)),
         Proportion = NewArea/OldArea,
         NewPop = totalpopE*Proportion) %>% 
  as.data.frame() %>% 
  group_by(SEN_DIST) %>% 
  summarize(totalpopE = sum(NewPop))

# identify FEMA risk pops by district and calculate pop pct
NFHZAbySenate <- vt_blkgrps_flood %>% 
  transmute(NewPop = NewPop, OldArea = as.numeric(st_area(.))) %>% 
  st_intersection(
    st_make_valid(
      st_transform(senate_districts, st_crs(vt_blkgrps_flood))
      )
    ) %>% 
  mutate(NewArea = as.numeric(st_area(.)),
         Proportion = NewArea/OldArea,
         nfhzaPop = as.integer(NewPop*Proportion)) %>% 
  as.data.frame() %>% 
  group_by(SEN_DIST) %>% 
  summarize(nfhzaPop = sum(nfhzaPop)) %>% 
  left_join(senate_names_pops, ., by = "SEN_DIST") %>% 
  replace_na(list(nfhzaPop = 0)) %>% 
  mutate(PctPopNFHZA = nfhzaPop/totalpopE*100)


# Create counts of block groups in flood zones by jurisdiction for popup 
muniNFHZA <- vt_blkgrps_sf %>%
  as.data.frame() %>%
  filter(floodRisk == "F" & POPSlabel != "No Pops of Concern") %>%
  group_by(TOWN) %>%
  summarize(BGsNFHZA = n())

houseNFHZA <- vt_blkgrps_sf %>%
  as.data.frame() %>%
  filter(floodRisk == "F" & POPSlabel != "No Pops of Concern") %>%
  group_by(REP_DIST) %>%
  summarize(BGsNFHZA = n())

senateNFHZA <- vt_blkgrps_sf %>%
  as.data.frame() %>%
  filter(floodRisk == "F" & POPSlabel != "No Pops of Concern") %>%
  group_by(SEN_DIST) %>%
  summarize(BGsNFHZA = n())


# simplify layers for faster loading and join BG stats
vt_towns <- vt_towns %>% 
  left_join(., muniNFHZA, by = c("NAME" = "TOWN")) %>%
  replace_na(list(BGsNFHZA = 0)) %>%
  ms_simplify(., keep = 0.1, keep_shapes = TRUE)

house_districts <- house_districts %>% 
  left_join(., houseNFHZA, by = "REP_DIST") %>%
  replace_na(list(BGsNFHZA = 0)) %>%
  ms_simplify(., keep = 0.1, keep_shapes = TRUE)

senate_districts <- senate_districts %>% 
  left_join(., senateNFHZA, by = "SEN_DIST") %>%
  replace_na(list(BGsNFHZA = 0)) %>%
  ms_simplify(., keep = 0.1, keep_shapes = TRUE)

# create layer of vt_blkgrps_sf for flood
# vt_blkgrps_sf_NFHZA <- vt_blkgrps_sf %>% 
#   filter(floodRisk == "F" & POPSlabel != "No Pops of Concern") %>% 
#   select(NAME, REP_DIST, SEN_DIST, TOWN, POPSlabel, floodtypes) %>% 
#   ms_simplify(., keep = 0.1, keep_shapes = TRUE)

vt_blkgrps_sf_FLOOD <- vt_blkgrps_sf %>% 
  filter((nfhzaRisk == "F" & POPSlabel != "No Pops of Concern") | 
           (rcRisk == "RC" & POPSlabel != "No Pops of Concern")) %>% 
  mutate(floodLabel = str_trim( #get rid of white space at begin & end
    gsub("^,*|(?<=,),|,$", "", # get rid of extra commas
         str_remove_all( # get rid of NAs
           paste(floodtypes, floodtype, sep = ","),
           pattern = "NA"),
         perl=T)
    )) %>% 
  select(NAME, REP_DIST, SEN_DIST, TOWN, POPSlabel, floodLabel) 



# # clean up
# rm(list = ls(pattern = paste("ne_","muni", "hevac", "nfhza", "_tracts_",
#                              "houseN", "senateN", "houseH", "senateH",
#                              "vt_blkgrps_sf$", "_list",sep = "|")))
```
---
# Evacuation
Flood and fluvial erosion are significant risks in Vermont. In the event of extensive flooding, evacuation may be required. For individuals and households with limited mobility, either due to inadequate access to transportation options or because of physical limitations, evacuation presents heightened risk. Evacuation may also prove especially difficult for individuals and households due to limited economic resources, difficulty understanding or accessing information, or low trust in official sources of information. 

* Over 61,000 people in Vermont live in a FEMA flood risk zone or river corridor subject to fluvial erosion. 
* Low income persons, households without a car, and people with disabilities are more likely to live in flood zones or river corridors compared to the general population. 

These interactive figures identify communities across Vermont that are most most vulnerable to the risks of inland flooding or coastal storm surge.

<br>

## Flood Hazard Exposure & Priority Populations by Census block group {.tabset}

### Flood
```{r mapNFHZA, fig.align="left", fig.cap="Map of Census block groups with the highest concentrations of one or more priority populations living in FEMA flood zones or river corridors."}
# create simplified towns layer for faster mapping
# download state outline
vt_state <- states(cb = TRUE) %>% 
  filter(NAME == "Vermont")
# filter out extra polygons, clip out boundaries from water, simplify
vt_towns_simple <- vt_towns %>% 
  filter(NAME != "County subdivisions not defined") %>% 
  ms_clip(., vt_state) %>% 
  ms_simplify(., keep = 0.1, keep_shapes = TRUE)

PopupHouse <- paste0("Vermont state House District ", "<b>",house_districts$REP_DIST,"</b>", " has ", "<b>",house_districts$BGsNFHZA,"</b>", " <b>Block Groups</b> with high percentages of priority populations living within FEMA flood zones or River Corridors.")

PopupSenate <- paste0("Vermont state Senate District ", "<b>",senate_districts$SEN_DIST,"</b>", " has ", "<b>",senate_districts$BGsNFHZA,"</b>", " <b>Block Groups</b> with high percentages of priority populations living within FEMA flood zones or River Corridors.")

Popup <- paste0(vt_blkgrps_sf_FLOOD$NAME, "<br/>",
                "<b>State House District:</b> ", vt_blkgrps_sf_FLOOD$REP_DIST, "<br/>",
                "<b>State Senate District:</b> ", vt_blkgrps_sf_FLOOD$SEN_DIST, "<br/>",
                "<b>Town:</b> ", vt_blkgrps_sf_FLOOD$TOWN, "<br/>",
                "<b>Priority Populations: </b>", vt_blkgrps_sf_FLOOD$POPSlabel, "<br/>",
                "<b>Flood risks: </b> ", vt_blkgrps_sf_FLOOD$floodLabel)

leaflet(width = "100%") %>% 
  addProviderTiles(providers$Stamen.TonerLite) %>% 
  addPolygons(data = vt_towns_simple,
              weight = 0.7,
              opacity = 1,
              color = "gray",
              fillOpacity = 0,
              label=~NAME, popup=~NAMELSAD, group='muni') %>% 
  addPolygons(data = house_districts,
              weight = 2,
              opacity = 1,
              color = "blue",
              dashArray = 3,
              fillOpacity = 0,
              # fillColor = "blue",
              label = ~REP_DIST,
              popup = PopupHouse,
              highlight = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 0,
                bringToFront = TRUE),
              group = "State House Districts") %>% 
  addPolygons(data = senate_districts,
              # fillColor = "red",
              weight = 2,
              opacity = 1,
              color = "green",
              dashArray = 3,
              fillOpacity = 0,
              label=~SEN_DIST,
              popup = PopupSenate,
              highlight = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 0,
                bringToFront = TRUE),
              group = "State Senate Districts") %>% 
  addPolygons(data = vt_blkgrps_sf_FLOOD, 
              color = "red",
              weight = 0.5,
              opacity = 0.7,
              # color = "white",
              dashArray = 3,
              fillOpacity = 0.7,
              label=~TOWN, 
              popup = Popup,
              highlight = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 0.7,
                bringToFront = TRUE)) %>% 
  addLegend(colors = "red",
            labels = "Flood Risk + Hi Priority Pops",
            position = "bottomleft") %>% 
  setView(lng = -72.7, 43.7, zoom = 7) %>% 
  # addMiniMap() %>% 
  addScaleBar(position = "bottomright") %>% 
  addSearchFeatures(targetGroups  = 'muni', 
                    options = searchFeaturesOptions(zoom=14, openPopup=TRUE, hideMarkerOnCollapse=T)) %>% 
  addLayersControl(
    overlayGroups = c("State House Districts","State Senate Districts"),
    options = layersControlOptions(collapsed = TRUE)
  ) %>% 
  hideGroup(c("State House Districts","State Senate Districts"))
```

<br>

<br>

### About the map

This map shows communities (i.e. Census Block Groups) with high percentages of one or more priority population groups (80th percentile for the state) *AND* that are living within FEMA flood risk zones *OR* River Corridors. 

The analysis of FEMA flood exposure presented here is based on the Federal Emergency Management Agency’s (FEMA) [National Flood Hazard Layer (NFHL)](https://www.fema.gov/national-flood-hazard-layer-nfhl), a digital version of FEMA’s most recent flood maps. Flood risk areas are subject to floods with an Annual Exceedance Probability (AEP) of 1% (also known as a ‘100-year’ flood) and areas subject to 0.2% AEP (also known as a ‘500-year’ flood). Areas within the 1% AEP flood zone are designated by FEMA as Special Flood Hazard Areas, and development within those zones must be covered by flood insurance. Areas within the 0.2% AEP are not currently regulated, but these areas are nevertheless subject to flood risk under more extreme, albeit less frequent, flooding circumstances. 

*Note that digital data on flood risk for much of northern Vermont is not currently available from FEMA. FEMA flood risk for northern Vermont is likely to be underestimated.*

The analysis of exposure to River Corridors is based on maps produced by the [Vermont Agency of Natural Resources (ANR)](https://floodready.vermont.gov/flood_protection/river_corridors_floodplains/river_corridors). The ANR’s river corridors represent areas surrounding a river that allow for the meandering, floodplain, and the riparian functions necessary to restore and maintain the naturally stable or least erosive form of a river in order to minimize erosion hazards over time. Lands within and immediately abutting a river corridor are at higher risk to fluvial erosion. Damage surveys in Vermont have shown that fluvial erosion, not inundation, is the most common
natural hazard type in Vermont.

Priority populations represent demographic groups that environmental justice policy and research have identified as being especially vulnerable to environmental burdens as a consequence of social or economic disadvantage, physical vulnerability, or historic and persistent discrimination and inequality. These include:

* People of color (i.e., persons who are of Hispanic ethnicity or racially not White)
* Low income persons (i.e., income less than 200% of the poverty line)
* Limited English speaking households (i.e., households where no adult speaks English "very well")
* Adults 25 years or older without a high school diploma
* Children under the age of 5
* Adults over the age of 64
* Individuals under the age of 18
* Adults 18 years or older with a physical disability
* Households without access to a car


Identifying populations at risk:

The highlighted areas at risk in the maps represent entire Census block groups. Please note, however, that not all of the Census block group is necessarily at risk of flooding or river corridors. The analysis identified only those *developed or occupied portions of block groups* falling within flood zones or river corridors based on ancillary data from the [National Land Cover Database (NLCD)](https://www.mrlc.gov/data/nlcd-2016-land-cover-conus). The NLCD is a nationwide geospatial database on land cover and land cover change at a 30m resolution produced by the US Geological Survey (USGS) in partnership with several federal agencies as part of the Multi-Resolution Land Characteristics Consortium (MRLC). Entire Census block groups are presented in these maps for ease of visualization at the scale of the state. Please consult the supporting Technical Report for this analysis for more detailed figures and tables. 

<br>

<br>

## Population-weighted flood risk exposure for priority populations {.tabset}

### Flood

```{r graphFEMA, fig.align="center", fig.cap="Percentage of priority populations living within FEMA flood zones or River Corridors."}
# Add status variable and recode demographic names
vt_FloodPops_df %>% 
  mutate(
    Status = case_when(
           Group == "Total Pop" ~ "State avg",
           PctFloodPop < vt_FloodPops_df[vt_FloodPops_df$Group=="Total Pop",10] ~ "Below state avg",
           PctFloodPop > vt_FloodPops_df[vt_FloodPops_df$Group=="Total Pop",10] ~ "Above state avg"),
    Group = recode(Group, "Minority" = "People of Color",
                        "No HS Dip" = "No HS Diploma"),
    PctFloodPop = round(PctFloodPop,1)) %>% 
  mutate(Group = factor(Group) %>% fct_reorder(PctFloodPop, .desc = TRUE)) %>% 
  mutate(group_index = as.numeric(Group)) %>% 
  hchart(., "bar", hcaes(x = group_index, y = PctFloodPop, 
                         group = Status, name = Group), 
                         color = c("#F7A35C", "#7CB5EC", "#000000"),
         pointWidth = 15) %>% 
  hc_yAxis(title = list(text = "Percentage of population"),
           labels = list(format = "{value}%")) %>% 
  hc_xAxis(title = NULL, type = "category", labels = list(step = 1)) %>% 
  hc_tooltip(pointFormat = "{series.name}: {point.y}%") %>% 
  hc_title(text = "Vermont Populations Living within Flood Zones or River Corridors", useHTML = TRUE)
```

<br>

<br>

### About the graph

This graph shows population-weighted flood risk or river corridor exposure by group:

* The analysis of FEMA flood exposure presented here is based on the Federal Emergency Management Agency’s (FEMA) [National Flood Hazard Layer (NFHL)](https://www.fema.gov/national-flood-hazard-layer-nfhl), a digital version of FEMA’s most recent flood maps. Flood risk areas are subject to floods with an Annual Exceedance Probability (AEP) of 1% (also known as a ‘100-year’ flood) and areas subject to 0.2% AEP (also known as a ‘500-year’ flood). Areas within the 1% AEP flood zone are designated by FEMA as Special Flood Hazard Areas, and development within those zones must be covered by flood insurance. Areas within the 0.2% AEP are not currently regulated, but these areas are nevertheless subject to flood risk under more extreme, albeit less frequent, flooding circumstances. 

*Note that digital data on flood risk for much of northern Vermont is not currently available from FEMA. FEMA flood risk for northern Vermont is likely to be underestimated.*

* The analysis of exposure to River Corridors is based on maps produced by the [Vermont Agency of Natural Resources (ANR)](https://floodready.vermont.gov/flood_protection/river_corridors_floodplains/river_corridors). The ANR’s river corridors represent areas surrounding a river that allow for the meandering, floodplain, and the riparian functions necessary to restore and maintain the naturally stable or least erosive form of a river in order to minimize erosion hazards over time. Lands within and immediately abutting a river corridor are at higher risk to fluvial erosion. Damage surveys in Vermont have shown that fluvial erosion, not inundation, is the most common
natural hazard type in Vermont.

Priority populations represent demographic groups that environmental justice policy and research have identified as being especially vulnerable to environmental burdens as a consequence of social or economic disadvantage, physical vulnerability, or historic and persistent discrimination and inequality. These include:

* People of color (i.e., persons who are of Hispanic ethnicity or racially not White)
* Low income persons (i.e., income less than 200% of the poverty line)
* Limited English speaking households (i.e., households where no adult speaks English "very well")
* Adults 25 years or older without a high school diploma
* Children under the age of 5
* Adults over the age of 64
* Individuals under the age of 18
* Adults 18 years or older with a physical disability
* Households without access to a car

Identifying populations at risk:

In order to identify counts and percentages of populations at risk, polygons of flood or river corridor risk zones were spatially intersected with the *developed or occupied portions of block groups* based on ancillary data from the [National Land Cover Database (NLCD)](https://www.mrlc.gov/data/nlcd-2016-land-cover-conus). The NLCD is a nationwide geospatial database on land cover and land cover change at a 30m resolution produced by the US Geological Survey (USGS) in partnership with several federal agencies as part of the Multi-Resolution Land Characteristics Consortium (MRLC). Populations at risk were calculated based on a process of areal apportionment. The population at risk from flooding was calculated as the product of the areal proportion of the intersecting flood and developed Block Group polygons: 
*Population at risk = Proportion of developed Block Group Intersection x Population of developed Block Group*
For example, if 10% of the developed area of a Census Block Group  intersected/overlapped with a flood polygon, it was assumed that 10% of the population is exposed to that flood risk. Assuming a population of 100 people in the developed portion of the Block Group, this would mean 100 x .10 = 10 people would be subject to flood risk.

<br>

<br>

## Flood Evacuation Burdens by Jurisdiction {.tabset}

### By municipality

```{r townTable, fig.align="center"}
# create object to hold complex headers for table
sketch1 = htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'City/Town'),
      th(align="center", colspan = 2, 'Flood Risk')
    ),
    tr(
      lapply(rep(c('Pop', 'Pct'), 1), th)
    )
  )
))

town_nfhzaPop_pct %>% 
  mutate(PctPopNFHZA = PctPopNFHZA/100) %>% 
  select(-totalpopE) %>% 
  arrange(NAMELSAD) %>% 
  datatable(., rownames = FALSE, options = list(pageLength = 10), 
            container = sketch1) %>% 
  formatRound(., columns = 2, digits = 0, mark = ",") %>% 
  formatPercentage(., columns = 3, digits = 1)
```

<!-- > Number of Census block groups with three or more cumualtive environmental burdens and high concentrations of populations of concern -->
<br>

<br>

### By state house district

```{r houseTable, fig.align="center"}
# create object to hold complex headers for table
sketch2 = htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'House District'),
      th(align="center", colspan = 2, 'Flood Risk')
    ),
    tr(
      lapply(rep(c('Pop', 'Pct'), 1), th)
    )
  )
))

NFHZAbyHouse %>% 
  select(-totalpopE) %>% 
  mutate(PctPopNFHZA = PctPopNFHZA/100) %>%  
  arrange(REP_DIST) %>% 
  datatable(., rownames = FALSE, options = list(pageLength = 10), 
            container = sketch2) %>% 
  formatRound(., columns = 2, digits = 0, mark = ",") %>% 
  formatPercentage(., columns = 3, digits = 1)
```

<br>

<br>

### By state senate district

```{r senateTable, fig.align="center"}
# create object to hold complex headers for table
sketch3 = htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'Senate District'),
      th(align="center", colspan = 2, 'Flood Risk')
    ),
    tr(
      lapply(rep(c('Pop', 'Pct'), 1), th)
    )
  )
))

NFHZAbySenate %>% 
  select(-totalpopE) %>% 
  mutate(PctPopNFHZA = PctPopNFHZA/100) %>%  
  arrange(SEN_DIST) %>% 
  datatable(., rownames = FALSE, options = list(pageLength = 10), 
            container = sketch3) %>% 
  formatRound(., columns = 2, digits = 0, mark = ",") %>% 
  formatPercentage(., columns = 3, digits = 1)
```

<br>

<br>

### About the tables

These tables show the number and percentage of the general population by jurisdiction exposed to FEMA flood risk or river corridors. 

* The analysis of FEMA flood exposure presented here is based on the Federal Emergency Management Agency’s (FEMA) [National Flood Hazard Layer (NFHL)](https://www.fema.gov/national-flood-hazard-layer-nfhl), a digital version of FEMA’s most recent flood maps. Flood risk areas are subject to floods with an Annual Exceedance Probability (AEP) of 1% (also known as a ‘100-year’ flood) and areas subject to 0.2% AEP (also known as a ‘500-year’ flood). Areas within the 1% AEP flood zone are designated by FEMA as Special Flood Hazard Areas, and development within those zones must be covered by flood insurance. Areas within the 0.2% AEP are not currently regulated, but these areas are nevertheless subject to flood risk under more extreme, albeit less frequent, flooding circumstances. 

*Note that digital data on flood risk for much of northern Vermont is not currently available from FEMA. FEMA flood risk for northern Vermont is likely to be underestimated.*

* The analysis of exposure to River Corridors is based on maps produced by the [Vermont Agency of Natural Resources (ANR)](https://floodready.vermont.gov/flood_protection/river_corridors_floodplains/river_corridors). The ANR’s river corridors represent areas surrounding a river that allow for the meandering, floodplain, and the riparian functions necessary to restore and maintain the naturally stable or least erosive form of a river in order to minimize erosion hazards over time. Lands within and immediately abutting a river corridor are at higher risk to fluvial erosion. Damage surveys in Vermont have shown that fluvial erosion, not inundation, is the most common
natural hazard type in Vermont.

Identifying populations at risk:

In order to identify counts and percentages of populations at risk, polygons of flood or river corridor risk zones were spatially intersected with the *developed or occupied portions of block groups* based on ancillary data from the [National Land Cover Database (NLCD)](https://www.mrlc.gov/data/nlcd-2016-land-cover-conus). The NLCD is a nationwide geospatial database on land cover and land cover change at a 30m resolution produced by the US Geological Survey (USGS) in partnership with several federal agencies as part of the Multi-Resolution Land Characteristics Consortium (MRLC). Populations at risk were calculated based on a process of areal apportionment. The population at risk from flooding was calculated as the product of the areal proportion of the intersecting flood and developed Block Group polygons: 
*Population at risk = Proportion of developed Block Group Intersection x Population of developed Block Group*
For example, if 10% of the developed area of a Census Block Group  intersected/overlapped with a flood polygon, it was assumed that 10% of the population is exposed to that flood risk. Assuming a population of 100 people in the developed portion of the Block Group, this would mean 100 x .10 = 10 people would be subject to flood risk.